---
title: "Detailed Dive: Variance Formulas and lambda-star"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Detailed Dive: Variance Formulas and lambda-star}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pppower)
```

## Mean estimator and `PPI++`

For labeled size $n$ and unlabeled size $N$:

*Estimator*
$$
\hat\theta_\lambda = \bar{Y}_L + \lambda(\bar{f}_U - \bar{f}_L).
$$

*Variance*
$$
\operatorname{Var}(\hat\theta_\lambda) =
\frac{\sigma_Y^2}{n}
 + \lambda^2 \sigma_f^2 \left(\frac{1}{N} + \frac{1}{n}\right)
 - \frac{2\lambda\,\operatorname{Cov}(Y,f)}{n}.
$$

Minimizing yields
$$
\lambda^* = \frac{\operatorname{Cov}(Y,f)}{(1 + n/N)\sigma_f^2}, \qquad
\operatorname{Var}(\hat\theta_{\lambda^*}) =
\frac{\sigma_Y^2}{n} - \frac{\operatorname{Cov}(Y,f)^2}{\sigma_f^2}\frac{N}{n(n+N)}.
$$

```{r lambda-star}
lambda_opt <- function(n, N, cov_y_f, var_f) cov_y_f / ((1 + n / N) * var_f)
lambda_opt(n = 200, N = 5000, cov_y_f = 0.6, var_f = 0.6)
```

## Power and required $n$

Given effect size $\Delta$, target power $1-\beta$, and level $\alpha$,
$$
\frac{\lvert\Delta\rvert}{\sqrt{\operatorname{Var}(\hat\theta_{\lambda^*})}}
\ge z_{1-\alpha/2} + z_{1-\beta}.
$$
`power_ppi_mean()` solves this directly; use it instead of hand algebra:

```{r n-required}
power_ppi_mean(
  delta      = 0.2,
  N          = 5000,
  n          = NULL,
  power      = 0.9,
  sigma_y2   = 1.0,
  sigma_f2   = 0.6,
  cov_y_f    = 0.6,
  lambda_mode = "oracle"
)
```

## Metrics-to-moments mapping (binary)

For sensitivity/specificity $(\text{sens}, \text{spec})$ and prevalence $p$,
$$
p_{\hat{Y}} = \text{sens}\,p + (1-\text{spec})(1-p), \qquad
\operatorname{Cov}(Y,f) = p(1-p)(\text{sens} + \text{spec} - 1).
$$
The helper `binary_moments_from_sens_spec()` returns $\operatorname{Var}(Y)$, $\operatorname{Var}(f)$, and
$\operatorname{Cov}(Y,f)$:

```{r sens-spec}
binary_moments_from_sens_spec(p = 0.3, sens = 0.85, spec = 0.85)
```

Use these outputs directly in `power_ppi_mean()`.
