<!-- README.md is generated from README.Rmd. Please edit that file -->

Website: <https://yiqunchen.github.io/pppower/>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pppower

**pppower** is an R package for power analysis and sample-size
calculation under the Prediction-Powered Inference (PPI/`PPI++`) framework.

## Prevalence Estimation with PPI

For binary outcomes (prevalence), you can estimate required **labeled**
sample size while leveraging model predictions on a large unlabeled set.
This is exactly what PPI is for:

- Inputs: target effect size \(\Delta\), unlabeled size \(N\), and model quality
  (e.g., sensitivity/specificity)
- Output: required labeled \(n\) for your target power
- Benefit: better predictions can reduce labeled annotation burden

Use `power_ppi_mean(..., n = NULL)` for one-sample prevalence and
`power_ppi_ttest_binary()` for two-group prevalence differences.

## Interactive Calculator

<div style="margin: 1rem 0 1.4rem; padding: 1.0rem; border: 2px solid #d9e7df; border-radius: 12px; text-align: center; background: #f8fbf9;">
  <a href="https://yiqunchen.github.io/pppower/articles/sample-size-calculator.html" style="display: inline-block; padding: 0.8rem 1.1rem; font-size: 1.05rem; font-weight: 700; text-decoration: none; border-radius: 10px; background: #155f54; color: #ffffff;">
    Open Interactive PPI Sample Size Calculator
  </a>
  <div style="margin-top: 0.55rem; font-size: 0.92rem; color: #3f4d4a;">
    Direct link to the live calculator page.
  </div>
</div>

## What is Prediction-Powered Inference?

Many modern studies have access to:

- **Labeled data** $(X_i, Y_i)$ for $i = 1, \ldots, n$ — expensive to
  obtain (human annotations, lab assays, expert reviews)
- **Unlabeled data** $\tilde{X}_j$ for $j = 1, \ldots, N$ — cheap and
  abundant, with ML predictions $f(\tilde{X}_j)$ available

PPI combines both sources for valid, efficient inference.  The `PPI++`
estimator for the population mean $\theta^* = E[Y]$ is:

$$\hat{\theta}_\lambda = \bar{Y}_L + \lambda(\bar{f}_U - \bar{f}_L)$$

where $\lambda$ controls the blend between labeled data and predictions.
The key insight: **good predictions reduce the amount of labeled data
needed** for a given level of statistical power.

## Key Formulas

The `PPI++` variance under the oracle $\lambda^*$ is:

$$\text{Var}(\hat{\theta}_{\lambda^*}) = \frac{\sigma_Y^2}{n} - \frac{\text{Cov}(Y,f)^2}{\sigma_f^2} \cdot \frac{N}{n(n+N)}$$

Solving for the required labeled sample size (quadratic in $n$):

$$n \geq \frac{\sigma_Y^2 - S^2 N + \sqrt{(\sigma_Y^2 - S^2 N)^2 + 4 S^2 N \sigma_Y^2 (1 - \rho_{Yf}^2)}}{2S^2}$$

where $S^2 = (\Delta / (z_{1-\alpha/2} + z_{1-\beta}))^2$.

**Rule of thumb:** The sample size ratio satisfies
$n_{\texttt{PPI++}} / n_{\text{classical}} \approx 1 - R^2$, where $R^2$
is the squared correlation between $Y$ and $f(X)$.  A predictor
explaining 80% of the variance cuts labeled data needs by ~80%.

## Supported Designs

| Design | Power / Sample Size | Simulation |
|--------|---------------------|------------|
| One-sample mean (continuous) | `power_ppi_mean()` | `simulate_ppi_mean()` |
| One-sample mean (binary) | `power_ppi_mean()` | `simulate_ppi_vanilla_mean()` |
| Two-sample $t$-test | `power_ppi_ttest()` | `simulate_ppi_ttest_binary()` |
| Paired $t$-test | `power_ppi_paired()` | — |
| 2x2 contingency table | `power_ppi_2x2()` | — |
| Regression contrast | `power_ppi_regression()` | — |

## Installation

You can install the development version of `pppower` from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yiqunchen/pppower")
```

## Quick Example

A three-step workflow: compute power → solve for required $n$ → verify
with Monte Carlo.

```{r quick-example, eval = FALSE}
library(pppower)

# Step 1: Compute power for a given sample size
power_ppi_mean(
  delta    = 0.2,
  N        = 5000,
  n        = 200,
  var_f    = 0.6,
  var_res  = 0.4,
  cov_y_f  = 0.6,
  alpha    = 0.05
)

# Step 2: Solve for required labeled n to achieve 80% power
power_ppi_mean(
  delta       = 0.2,
  N           = 5000,
  n           = NULL,
  power       = 0.80,
  sigma_y2    = 1.0,
  sigma_f2    = 0.49,
  cov_y_f     = 0.63,
  lambda_mode = "oracle"
)

# Step 3: Verify with Monte Carlo simulation
simulate_ppi_mean(
  R     = 2000,
  n     = 100,
  N     = 5000,
  delta = 0.2,
  var_f   = 0.49,
  var_res = 0.51,
  alpha = 0.05
)
```

## Input Modes

`pppower` supports three ways to specify variance components:

**1. Direct moments** — supply $\sigma_Y^2$, $\sigma_f^2$, and
$\text{Cov}(Y, f)$ directly:

```r
power_ppi_mean(delta = 0.2, N = 5000, n = 200,
                  sigma_y2 = 1.0, sigma_f2 = 0.49, cov_y_f = 0.63)
```

**2. Prediction variance / residual variance** — decompose outcome
variance via `var_f` and `var_res`:

```r
power_ppi_mean(delta = 0.2, N = 5000, n = 200,
                  var_f = 0.49, var_res = 0.51, cov_y_f = 0.49)
```

**3. Metrics interface** — supply sensitivity/specificity (binary) or
MSE/$R^2$ (continuous):

```r
power_ppi_mean(delta = 0.05, N = 5000, n = 200,
                  metrics = list(sensitivity = 0.85, specificity = 0.90,
                                 p_y = 0.3, m_obs = 200),
                  metric_type = "classification")
```

## Vignettes

- **Quickstart: Prediction-Powered Power & Sample Size** — end-to-end
  walkthrough of power curves and Type I error checks
  (`vignette("intro-ppi")`)
- **Detailed Dive: Sample Size & Metrics Interface** — how
  `power_ppi_mean(..., n = NULL)` works across input modes, including
  regression contrasts (`vignette("ppi-sample-size")`)
- **Detailed Dive: Variance Formulas and lambda-star** — mathematical
  derivations behind the `PPI++` variance and oracle $\lambda^*$
  (`vignette("deep-dive-math")`)
- **2x2 Contingency Table Calculator** — odds-ratio and relative-risk
  planning with binary surrogates (`vignette("calculator-2x2")`)

## Citation

If you use `pppower` in your research, please cite:

```
@software{pppower,
  title = {pppower: Prediction-Powered Inference Power Calculations},
  author = {Guo, Moran and Chen, Yiqun},
  url = {https://github.com/yiqunchen/pppower},
  year = {2025}
}
```
